---
output:
  pdf_document: default
  html_document: default
---
title: "Practical Machine Learning Project"
author: "Mohamed Osama"
date: "7/25/2020"
---

# Practical Machine Learning Project : Prediction Assignment 

## 1. Overview

This document is the final report of the Peer Assessment project from Coursera’s course Practical Machine Learning, as part of the Specialization in Data Science. It was built up in RStudio, using its knitr functions, meant to be published in html format.
This analysis meant to be the basis for the course quiz and a prediction assignment writeup. The main goal of the project is to predict the manner in which 6 participants performed some exercise as described below. This is the “classe” variable in the training set. The machine learning algorithm described here is applied to the 20 test cases available in the test data and the predictions are submitted in appropriate format to the Course Project Prediction Quiz for automated grading.




## 2. Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

Read more: http://groupware.les.inf.puc-rio.br/har#ixzz3xsbS5bVX



## 3. Data Loading and Exploratory Data Analysis

### a) Dataset Overview
The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from http://groupware.les.inf.puc-rio.br/har. Full source:

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. “Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human ’13)”. Stuttgart, Germany: ACM SIGCHI, 2013.

My special thanks to the above mentioned authors for being so generous in allowing their data to be used for this kind of assignment.

A short description of the datasets content from the authors’ website:

“Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes. Participants were supervised by an experienced weight lifter to make sure the execution complied to the manner they were supposed to simulate. The exercises were performed by six male participants aged between 20-28 years, with little weight lifting experience. We made sure that all participants could easily simulate the mistakes in a safe and controlled manner by using a relatively light dumbbell (1.25kg)."

### b) Environment Preparation    
We first upload the R libraries that are necessary for the complete analysis.   


```{r}
rm(list=ls())                # free up memory for the download of the data sets



library(caret)
library(rpart)
library(rpart.plot)
library(rattle)
library(randomForest)
library(corrplot)
set.seed(12345)

```



### c) Data Loading and Cleaning   

The next step is loading the dataset from the URL provided above. The training dataset is then partinioned in 2 to create a Training set (70% of the data) for the modeling process and a Test set (with the remaining 30%) for the validations. The testing dataset is not changed and will only be used for the quiz results generation.    

```{r}
# set the URL for the download
UrlTrain <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
UrlTest  <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

# download the datasets
training <- read.csv(url(UrlTrain))
testing  <- read.csv(url(UrlTest))

# create a partition with the training dataset 
inTrain  <- createDataPartition(training$classe, p=0.7, list=FALSE)
TrainSet <- training[inTrain, ]
TestSet  <- training[-inTrain, ]
dim(TrainSet)
```

```{r}
dim(TestSet)
```

Both created datasets have 160 variables. Those variables have plenty of NA, that can be removed with the cleaning procedures below. The Near Zero variance (NZV) variables are also removed and the ID variables as well.   


```{r}
# remove variables with Nearly Zero Variance
NZV <- nearZeroVar(TrainSet)
TrainSet <- TrainSet[, -NZV]
TestSet  <- TestSet[, -NZV]
dim(TrainSet)

```

```{r}
dim(TestSet)
```
```{r}
# remove variables that are mostly NA
AllNA    <- sapply(TrainSet, function(x) mean(is.na(x))) > 0.95
TrainSet <- TrainSet[, AllNA==FALSE]
TestSet  <- TestSet[, AllNA==FALSE]
dim(TrainSet)

```

```{r}
dim(TestSet)
```

```{r}
# remove identification only variables (columns 1 to 5)
TrainSet <- TrainSet[, -(1:5)]
TestSet  <- TestSet[, -(1:5)]
dim(TrainSet)
```
```{r}
dim(TestSet)
```
With the cleaning process above, the number of variables for the analysis has been reduced to 54 only.    

### d) Correlation Analysis   

A correlation among variables is analysed before proceeding to the modeling procedures.    

```{r}
corMatrix <- cor(TrainSet[, -54])
corrplot(corMatrix, order = "FPC", method = "color", type = "lower", 
         tl.cex = 0.8, tl.col = rgb(0, 0, 0))
```

The highly correlated variables are shown in dark colors in the graph above. To make an evem more compact analysis, a PCA (Principal Components Analysis) could be performed as pre-processing step to the datasets. Nevertheless, as the correlations are quite few, this step will not be applied for this assignment.    
## 4. Prediction Model Building

  

### 4.1. Decision Tree Model    

```{r}
# model fit
set.seed(1813)
fit_decision_tree <- rpart(classe ~ ., data = TrainSet, method="class")
fancyRpartPlot(fit_decision_tree)
```

Predictions of the decision tree model on TestSet.
```{r}
# prediction on Test dataset
predict_decision_tree <- predict(fit_decision_tree, newdata = TestSet, type="class")
conf_matrix_decision_tree <- confusionMatrix(predict_decision_tree,factor(TestSet$classe))
conf_matrix_decision_tree
```
The predictive accuracy of the decision tree model is relatively low at 73.4 %.   
**Plot the predictive accuracy of the decision tree model.**     
```{r}
plot(conf_matrix_decision_tree$table, col = conf_matrix_decision_tree$byClass, 
     main = paste("Decision Tree Model: Predictive Accuracy =",
                  round(conf_matrix_decision_tree$overall['Accuracy'], 4)))
```

### 4.2. Generalized Boosted Model (GBM)   

```{r}
set.seed(1813)
ctrl_GBM <- trainControl(method = "repeatedcv", number = 5, repeats = 2)
fit_GBM  <- train(classe ~ ., data = TrainSet, method = "gbm",
                  trControl = ctrl_GBM, verbose = FALSE)
fit_GBM$finalModel
```


```{r}
predict_GBM <- predict(fit_GBM, newdata = TestSet)
conf_matrix_GBM <- confusionMatrix(predict_GBM, factor(TestSet$classe))
conf_matrix_GBM
```

**The predictive accuracy of the GBM is relatively high at 98.78 %.**     

### 4.3. Random Forest Model     
```{r}
set.seed(1813)
ctrl_RF <- trainControl(method = "repeatedcv", number = 5, repeats = 2)
fit_RF  <- train(classe ~ ., data = TrainSet, method = "rf",
                  trControl = ctrl_RF, verbose = FALSE)
fit_RF$finalModel
```


Predictions of the Random Forest model on TestSet.

```{r}
predict_RF <- predict(fit_RF, newdata = TestSet)
conf_matrix_RF <- confusionMatrix(predict_RF, factor(TestSet$classe))
conf_matrix_RF
```

**The predictive accuracy of the Random Forest model is excellent at 99.9 %.**   
## 5. Applying the Best Predictive Model to the Test Data     

To summarize, the predictive accuracy of the three models evaluated is as follows:

Decision Tree Model: 73.42 %       
Generalized Boosted Model: 98.87 %         
Random Forest Model: 99.92 %         


**The Random Forest model is selected and applied to make predictions on the 20 data points from the original testing dataset (data_quiz).**         

```{r}
predict_quiz <- predict(fit_RF, newdata = testing)
predict_quiz
```


